Applying Renormalization Group Theory to the Square Well Liquid

### 2014-06-05 papers/thesis-roth/project.tex
project report Tested on elliot. Took 0:20.40.

% special values for n
\newcommand{\npart}{\ensuremath{n_{part}}}
\newcommand{\nliq}{\ensuremath{n_{liquid}}}
\newcommand{\nvap}{\ensuremath{n_{vapor}}}

% Free energy terms
\newcommand{\Fattr}{\ensuremath{F_\text{attr}(T,n)}}

\newcommand{\fid}{\ensuremath{f_\text{ID}(T,n)}}
\newcommand{\fhs}{\ensuremath{f_\text{HS}(T,n)}}
\newcommand{\fdisp}{\ensuremath{f_\text{disp}(T,n)}}
\newcommand{\fattr}{\ensuremath{f_\text{attr}(T,n)}}

% lambda
\newcommand{\lambdaSW}{\ensuremath{\lambda_\text{SW}}}
\newcommand{\lambdaRG}{\ensuremath{\lambda_\text{RG}}}

% pair potential
\newcommand{\1}{\ensuremath{\textbf{r}_1}}
\newcommand{\2}{\ensuremath{\textbf{r}_2}}
\newcommand{\3}{\ensuremath{\textbf{r}_3}}
\newcommand{\4}{\ensuremath{\textbf{r}_4}}

% RG stuff
\newcommand{\FbarD}{\ensuremath{\bar{F}_D(T,n)}}
\newcommand{\UbarD}{\ensuremath{\bar{U}_D(T,n)}}

@@

Near the critical point of a liquid, there are density fluctuations at
all length scales. In this regime, mean field theories break down. \fixme{re-word the next sentence} As
the critical point is approached, there becomes less and less of a
difference in free energy of the liquid and vapor states; this causes
difficulty in modeling liquid-vapor coexistence. This work develops a
method for calculating liquid-vapor coexistence, and applies
renormalization group theory to the square well liquid.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Gasses have been accurately modeled since the 19$^{th}$
century\cite{Lederman92} and formulation of the kinetic theory of gasses. With the
advent of quantum mechanics in the early 20$^{th}$ century, we have
had a pretty good grasp on solids. A robust model of liquids continues
to be elusive. Liquids are not periodic structures like solids, but
neither are they disperse, as are gasses.

However, understanding liquids has both philosophical and economic
desireability. Liquid water is---as far as we know---essential for
developing and sustaining life. All celluar processes take place in an
aqueous environment. Understanding liquids is in many ways a step
towards understanding the development and existence of life. On the
economic side of the coin, we recognize liquids to be an important
part of a wide array of industrial applications.

One particularly troublesome regime is around the critical
point of a liquid. At the appropriate density and temperature, the
distinction between liquid and vapor dissapears. Beyond this point,
you have a ``supercritical fluid.'' Supercritical fluids are used in
many industrial processes.\cite{Perrut00} For example (and of personal
interest), supercritical fluids are used in coffee decaffeination and
the extraction of hops resins for brewing beer.

It is the aim of renormalization group theory to tackle the critical
point. The reason that calculations are dificult near the critical
point is that there are large-scale fluctuations of the density. In
fact, there are density fluctuations at \emph{all} length
scales, which is why mean field theories fail. Renormalization group theory is an iterative procedure that
was developed in th 1950's, and Kenneth Wilson's work with the theory
in the 1970's won him the 1982 Nobel Prize in physics.\cite{Nobel82}

\fixme{Transition here?}

The simplest model of a liquid is to consider each molecule to be a
hard sphere (i.e.~spheres cannot overlap) with a square-well potenial
between any pair of spheres. This is called the ``square well
liquid,'' and it is a widely used approximation to simple
liquids. This is one of the foundations of this work.

We use statistical mechanics to study liquids; this requires large
sample sizes, long simulation times, or some combination of the
two. Monte Carlo simulations are exact, if you allow them to run
``long enough.'' The Roundy group uses classical Density-funtional
Theory (DFT) to calculate free energies and probability distributions
of bulk liquids. These calculations are ultimately compared to Monte Carlo simulations.

The end goal of this work is to investigate into whether or not
implementing renormalization group theory is worthwhile to the Roundy
Research Group. Many other groups use renormalization group theory,
but none of them tell the iteration depth used in their
calculations.\cite{Forte11,Ramana12,Ghobadi13,Wang13,Zhao13} If only
few iterations are required, then this would certainly be an effective
tool to use. However, if many iterations are required, the
computational cost could be unwieldy.

In section~\ref{sec:theory}, I present some of the main theories I
deal with: renormalization group theory, thermodynamic perturbation
theory and the square well liquid free energy. In
section~\ref{sec:methods} I discuss the the methods used. I cover the
calculation of liquid-vapor coexistence curves in some detail, as
caluclating coexistence near the critical point is tricky. The
results---such as they are---are covered in section~\ref{sec:results}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}\label{sec:theory}

\subsection{Renormalization Group Theory}\label{subsec:RGT}
Renormalization group theory (RGT) is an iterative procedure that was
developed to deal with large-scale fluctuations of the order parameter
near the critical point of a system. The formulation used here is
provided by Forte~\textit{et al}.\cite{Forte11}

We start by dividing the total volume into cells of volume $V_i$. We
assume that within a cell the number density $n(\rr)$ varies only slightly, and that the
cells do not interact with one another. This comes from Wilson's
phase cell method.\cite{Ramana12}

Then consider a baseline free enrgy $F_0(T,n)$, usually based on a
mean field theory. We assume this baseline accounts for all
fluctuations of wavelength $\lambdaRG < \lambda_0$. We then
double the size of our phase cell and add a correction to the free
energy $\delta F_1(T,n)$. This correction accounts for fluctuations
with a wavelength $\lambdaRG = 2\lambda_0$. The new free
energy, $F_1(T,n)$, is
\begin{align}
  F_1(T,n) &= F_0(T,n) + \delta F_1(T,n)
\end{align}
Then double the length scale again; $\delta F_2(T,n)$ accounts for
fluctionations with $\lambdaRG = 4\lambda_0$ and
\begin{align}
  F_2(T,n) &= F_1(T,n) + \delta F_2(T,n)
\end{align}
As we continue this process, the $i^{th}$ correction accounts for
fluctuations with $\lambdaRG = 2^i\lambda_0$, and
\begin{align}
  F_i(T,n) &= F_{i-1}(T,n) + \delta F_i(T,n)
\end{align}
In the end, your free energy is
\begin{align}
  F(T,n) = F_0(T,n) + \sum_{i=1}^\infty\delta F_i(T,n) + \Fattr
\end{align}

Where $\Fattr$ is the contribution due to the longest-wavelength
density fluctuations. At this stage, I do not entirely understand why
Forte separates out $\Fattr$.

The $\delta F_i(T,n)$ term is a difference in free energies:
\begin{align}
  \delta F_i(T,n) &= -\kT\ln Z - \left( -\kT\ln Z^* \right) \nonumber \\
  &= -\kT\ln\left[ \frac{Z(T,n)}{Z^*(T,n)} \right]
\end{align}
Here, $Z$ is the partition function for the entire system, and $Z^*$
is the partition function of a reference system consisting only of the
term $\FbarD$. Forte gives these terms as
\begin{align}
  Z(T,n) &= \int dx\, \exp\left[ -\frac{V}{\kT}\left( \FbarD + \bar{U}(T,n \right) \right] \label{eqn:Z} \\
  Z^*(T,n) &= \int dx\, \exp\left[ -\frac{V}{\kT}\FbarD \right] \label{eqn:Zstar}
\end{align}
where $x$ is the amplitude of density fluctuations. $\FbarD$ and $\UbarD$ are formulated similar to averages:
\begin{align}
  \FbarD &= \frac{F_{i-1}(T,n+x) + F_{i-1}(T,n-x)}{2} - F_{i-1}(T,n) \\
  \UbarD &= \frac{U(T,n+x,\lambdaRG) + U(T,n-x,\lambdaRG)}{2} - U(T,n,\lambdaRG)
\end{align}
Finally, $U(T,n,\lambdaRG)$ is given by Forte as
\begin{align}
  U(T,n,\lambdaRG) = &{} g^\text{HS}(n_{eff};\sigma)\alpha(\lambdaRG) - \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)\alpha(\lambdaRG)\omega^2(\lambdaRG) \nonumber \\
  &{} + \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)^2\alpha(\lambdaRG)\gamma(\lambdaRG) \label{eqn:URG} % URG! It sounds like it hurt! :P
\end{align}
with
\begin{align}
  \alpha(\lambdaRG) &= \frac{2}{3}\pi\epsilon\sigma^3(\lambdaRG^3 - 1) \\
  \omega^2(\lambdaRG) &= \frac{1}{5}\sigma^2\frac{\lambdaRG^5 - 1}{\lambdaRG^3 - 1} \\
  \gamma(\lambdaRG) &= \frac{1}{70}\sigma^4\frac{\lambdaRG^7 - 1}{\lambdaRG^3 - 1}
\end{align}
The pair correlation function $g(n,r)$ tells us how likely we are to
find another hard sphere some distance away, given some density. In
equation~\ref{eqn:URG} we are considering this quantity at contact ($r
= \sigma$) with an effective density $n_{eff}$.

\fixme{Transition?}

@@

Thermodynamic perturbation theory (TPT) gives a good description of
simple liquids, as long as you are far from the critical point---which
is most of the time. TPT is a rigorous power series
expansion in a small parameter; as long as the parameter is small, the
theory is valid. This discussion follows primarily from \textit{Theory
  of Simple Liquids}, by Hansen and McDonald\cite{Hansen06}.

Thermodynamic perturbation theory treats the intermolecular potential
as separable into two parts, one for short-range repulsion and another
for long-range attraction.

The pair potential $v(\1,\2)$ describes the interaction between two
hard spheres. For the square-well model, the pair potential is
\begin{align}
  v_\text{HS}(\1,\2) &=
    \begin{cases}
      \infty & r < \sigma \\
      0 & \sigma < r
    \end{cases}
\end{align}

We consider this as our referecne, and take a square well attraction as a perturbation:
\begin{align}
  w_\text{SW}(\1,\2) &= - \varepsilon \text{ for } \sigma < r < \lambdaSW\sigma
\end{align}
so that we have
\begin{align}
  v_\lambda(\1,\2) &= v_\text{HS}(\1,\2) + \lambdaSW w_\text{SW}(\1,\2) \label{eqn:small-perturbation}
\end{align}

The total potential energy of our system and its derivatives with
resepect to $\lambdaSW$ (which become important later) are
\begin{align}
  \widetilde{V} &= \sum_{i=1}^N\sum_{j>i}^N v_\lambda(\mathbf{r}_i,\mathbf{r}_j) \\ % \label{eq:VN} \\
  \frac{\partial \widetilde{V}}{\partial\lambdaSW} &= \sum_{i=1}^N\sum_{j>i}^N w_\text{SW}(\mathbf{r}_i,\mathbf{r}_j) \nonumber \\
  &\equiv W_N \\
  \frac{\partial^2 \widetilde{V}}{\partial\lambdaSW^2} &= 0
\end{align}

The \emph{configuration integral} (the potential energy component of
the partition function) and its derivative with respect to $\lambdaSW$
are
\begin{align}
  Z_\lambda &= \int d\rr^N\, e^{-\beta \widetilde{V}} \\
  \frac{\partial Z_\lambda}{\partial\lambdaSW} &=  \int d\rr^N\, \frac{\partial}{\partial\lambdaSW}\left( e^{-\beta \widetilde{V}} \right) \nonumber \\
  &= \int d\rr^N\, -\beta\frac{\partial \widetilde{V}}{\partial\lambda}e^{-\beta \widetilde{V}} \nonumber \\
  &= \int d\rr^N\, -\beta W_N e^{-\beta \widetilde{V}}
\end{align}
And the non-ideal part of the free energy (i.e.~the excess free
energy) and its derivative are
\begin{align}
  F &= -\kT\ln\left( \frac{Z_\lambda}{V^N} \right) \\
  \frac{\partial F}{\partial\lambdaSW} &= \frac{V^N}{Z_\lambda}\frac{\partial}{\partial\lambdaSW}\left( \frac{Z_\lambda}{V^N} \right) \nonumber \\
  &= -\kT\left( \frac{1}{Z_\lambda}\frac{\partial Z_\lambda}{\partial\lambdaSW} \right) \nonumber \\
  &= \frac{1}{Z_\lambda} \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \label{eqn:dfdlambda_Z}\\
  &= \left\langle W_N \right\rangle_{\lambdaSW} \label{eqn:dfdlambda}
\end{align}
(where $-\kT$ and $-\beta$ have divided out to $1$ before equation~\ref{eqn:dfdlambda_Z}). Here, $V^N$ is
the total volume taken to the $N^{th}$ power, and $\left\langle W_N
\right\rangle_{\lambdaSW}$ is the ensemble average of $W_N$, with the
subscript indicating some value of $\lambdaSW$.

%% \fixme{The following could probably go someplace else (is it even necessary?):
%%   We can simplify this even further by noting
%%   \begin{align}
%%     \left\langle W_N \right\rangle &= \left\langle \sum_i\sum_jw(\mathbf{r}_i,\mathbf{r}_j) \right\rangle \\
%%     &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle
%%     \intertext{which gives us}
%%     \frac{\partial F}{\partial\lambda} &= \frac{1}{2}N^2\left\langle w(\1,\2) \right\rangle_\lambda
%% \end{align}}

Note that equation~\ref{eqn:dfdlambda} is exact; now take a series expansion:
\begin{widetext}
\begin{align}
  \left\langle W_N \right\rangle_{\lambdaSW} &= \left\langle W_N\right\rangle_{\lambdaSW = 0} + (\lambdaSW)\frac{\partial\left\langle W_N \right\rangle_{\lambdaSW}}{\partial\lambdaSW}\bigg|_{\lambdaSW = 0} + \mathcal{O}(\lambdaSW^2) \label{eqn:Wn-expansion}
\end{align}
%% \end{widetext}
The derivative term is
%% \begin{widetext}
  \begin{align}
    \frac{\partial\left\langle W_N \right\rangle_{\lambdaSW}}{\partial\lambdaSW} &= \frac{\partial}{\partial\lambdaSW}\left( \frac{1}{Z_\lambda}\int d\rr^N\, W_N e^{-\beta \widetilde{V}}\right) \nonumber \\
    &= \frac{1}{Z_\lambda}\frac{\partial}{\partial\lambdaSW}\left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right) - \left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\frac{1}{Z_\lambda^2}\frac{\partial}{\partial\lambdaSW}Z_\lambda \nonumber \\
    &= \frac{1}{Z_\lambda}\int d\rr^N\, \left( e^{-\beta \widetilde{V}}\frac{\partial}{\partial\lambdaSW}W_N + W_N\frac{\partial}{\partial\lambdaSW}e^{-\beta \widetilde{V}} \right) - \left( \int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\frac{1}{Z_\lambda^2}\int d\rr^N\, \frac{\partial}{\partial\lambdaSW}e^{-\beta \widetilde{V}}
  \end{align}
%% \end{widetext}

Noting that $\partial W_N/\partial\lambda = 0$, we have
%% \begin{widetext}
  \begin{align}
    \frac{\partial\left\langle W_N \right\rangle_{\lambdaSW}}{\partial\lambdaSW} &= \frac{1}{Z_\lambda}\int d\rr^N (-\beta) W_N^2 e^{-\beta \widetilde{V}} - \frac{1}{Z_\lambda^2}\left(\int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)\left( \int d\rr^N\, (-\beta)W_Ne^{-\beta \widetilde{V}} \right) \nonumber \\
    &= -\beta\left[ \frac{1}{Z_\lambda}\int d\rr^N\, W_N^2e^{-\beta \widetilde{V}} - \left( \frac{1}{Z_\lambda}\int d\rr^N\, W_N e^{-\beta \widetilde{V}} \right)^2 \right] \nonumber \\
    &= -\beta\left[ \left\langle W_N^2 \right\rangle_{\lambdaSW} - \left\langle W_N \right\rangle_{\lambdaSW}^2 \right] \nonumber \\
    &\equiv -\beta\sigma_{\lambdaSW}^2
  \end{align}
\end{widetext}

Combining this result with equation~\ref{eqn:Wn-expansion} gives us
\begin{align}
  \left\langle W_N \right\rangle_{\lambdaSW} &= \left\langle W_N\right\rangle_{\lambdaSW = 0} - \beta\lambda\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\lambdaSW^2) \label{eqn:Wn-expansion-simplified}
\end{align}
and we can insert this into equation~\ref{eqn:dfdlambda}:
\begin{align}
  \frac{\partial F}{\partial\lambdaSW} &= \left\langle W_N\right\rangle_{\lambdaSW = 0} - \beta\lambda\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\lambdaSW^2)
\end{align}

Now integrate this expression over $F\in[F_0,F]$ (\textsc{N.B.}~this is different than the RGT baseline free energy $F_0(T,n)$ in section~\ref{subsec:RGT}) and $\lambda\in[0,1]$ to obtain
\begin{align}
  \int_{F_0}^F\, dF &= \int_0^1 d\lambdaSW\, \left\langle W_N\right\rangle_{\lambdaSW = 0} - \beta\lambda\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\lambdaSW^2) \nonumber \\
  \intertext{\fixme{Do I need that first line?}}
  \hookrightarrow F - F_0 &= (1-0)\left\langle W_N \right\rangle_{\lambdaSW} - \frac{1}{2}\beta(1 - 0)^2\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\beta^2) \nonumber \\
  \hookrightarrow F &= F_0 + \left\langle W_N \right\rangle_{\lambdaSW} - \frac{1}{2}\beta\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\beta^2) \label{eqn:F-inTermsW}
\end{align}
Multiply by $\beta$ and we have the \emph{high-temperature expansion}, first derived by
Zwanzig:\cite{Zwanzig54}
\begin{align}
  \beta F &= \beta F_0 + \beta \left\langle W_N \right\rangle_{\lambdaSW=0} - \beta^2\lambda\sigma_{\lambdaSW=0}^2 + \mathcal{O}(\beta^3)
\end{align}

Zwanzig further showed that the $n^{th}$ term could be written in
terms of the mean fluctuations $\left\langle \left[ W_N - \left\langle
  W_N\right\rangle_{\lambdaSW=0} \right]^\nu \right\rangle_{\lambdaSW=0}$
with $\nu \leq n$. This requires us to know the form of higher-order
correlation functions of the reference system, which can be quite
complicated. In the next section we will see a simple aproximation for
the $n=2$ term.

>>>>

% special values for n
\newcommand{\npart}{\ensuremath{n_\textit{part}}}
\newcommand{\nliq}{\ensuremath{n_\textit{liquid}}}
\newcommand{\nvap}{\ensuremath{n_\textit{vapor}}}
\newcommand{\neff}{\ensuremath{n_\textit{eff}}}

% Free energy terms
\newcommand{\Fattr}{\ensuremath{F_\text{attr}(T,n)}}
\newcommand{\Fex}{\ensuremath{F_\text{ex}}}

\newcommand{\fid}{\ensuremath{f_\text{ID}(T,n)}}
\newcommand{\fhs}{\ensuremath{f_\text{HS}(T,n)}}
\newcommand{\fdisp}{\ensuremath{f_\text{disp}(T,n)}}
\newcommand{\fattr}{\ensuremath{f_\text{attr}(T,n)}}
\newcommand{\fsw}{\ensuremath{f_\text{SW}(T,n)}}

% lambda
\newcommand{\lambdaSW}{\ensuremath{\lambda_\text{sw}}}
\newcommand{\lambdaRG}{\ensuremath{\lambda_\text{rg}}}

% TPT
\newcommand{\III}{\ensuremath{\textbf{r}_{12}}} % Roman numerals for \12
\newcommand{\Vtilde}{\ensuremath{\widetilde{V}}}
\newcommand{\Wtilde}{\ensuremath{\widetilde{W}}}

% RG stuff
\newcommand{\FbarD}{\ensuremath{\bar{F}_D(T,n,x)}}
\newcommand{\UbarD}{\ensuremath{\bar{U}_D(T,n,x,\lambdaRG)}}

@@

This work develops a method for calculating liquid-vapor coexistence,
and applies renormalization group theory to the square well
liquid. Near the critical point of a liquid, there are density
fluctuations at all length scales. In this regime, mean field theories
break down. As the liquid approaches the critical point, the
difference in free energy between the liquid and vapor states
diminishes; this causes difficulty in modeling liquid-vapor
coexistence.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Gases have been accurately modeled since the 19$^{th}$
century\cite{Lederman92} and formulation of the kinetic theory of gasses. With the
advent of quantum mechanics in the early 20$^{th}$ century, we have
had a pretty good grasp on solids. A robust model of liquids continues
to be elusive. Liquids are not periodic structures like solids, but
neither are they disperse, as are gasses.

However, understanding liquids has both philosophical and economic
desireability. Liquid water is---as far as we know---essential for
developing and sustaining life. All cellular processes take place in an
aqueous environment. Understanding liquids is in many ways a step
towards understanding the development and existence of life. On the
economic side of the coin, we recognize liquids to be an important
part of a wide array of industrial applications.

One particularly troublesome regime is around the critical
point of a liquid. At the appropriate density and temperature, the
distinction between liquid and vapor disappears. Beyond this point,
you have a ``supercritical fluid.'' Supercritical fluids are used in
many industrial processes.\cite{Perrut00} For example (and of personal
interest), supercritical fluids are used in coffee decaffeination and
the extraction of hops resins for brewing beer.

In this work, I focus exclusively on what is called the square well
liquid. The simplest model of a liquid is to consider each molecule to
be a hard sphere (i.e.~spheres which cannot overlap) with a square-well
potential between any pair of spheres. This is a widely used
approximation to simple liquids, and one of the foundations of
this work.

There are several approaches to study liquids, each with advantages
and disadvantages. Some methods, such as Monte Carlo and Molecular
Dynamics, are simulations of systems with a discrete number of
particles. Other methods are more abstract in nature. Methods such as
Thermodynamic Perturbation Theory, Density Functional Theory and
Renormalization Group Theory are mathematical theories that were not
necessarily developed with computational methods in mind (though in
practice numerical solutions are required).

Monte Carlo simulations (MC) are one method. These simulations can
give exact answers, if you run them ``long enough.''  So, MC methods
are very accurate, but very computationally expensive. Molecular
dynamics simulations (MD) calculate forces on molecules over a
particular span of time. These simulations are also accurate but even
more computationally expensive than MC. Calculations simulating only a
few hundred molecules for picoseconds to nanoseconds of interactions
can be expected.\cite{Hughes13}

Thermodynamic Perturbation Theory (TPT) gives a good description of
liquid systems far from the critical point. Similar to quantum
perturbation theory, TPT utilizes the power and simplicity of a power
series expansion in a small parameter. Unfortunately, TPT does break
down near the critical point.

Density Functional Theory (DFT) implements calculations based on
ground-state electron density distributions. DFT is good for solid
state systems, but liquids are too complicated. Simple DFT MD
simulations ``predict'' a melting point of water at 120$^\circ$C. More
complicated---and therefore computationally expensive---models are
slightly more accurate; one model calculates a melting point of ice as
87$^\circ$C.\cite{Yoo11} Classical DFT (cDFT) describes a fluid by a
free energy functional of the particle density
distribution.\cite{Krebs13} cDFT efficiently calculates density
profiles for inhomogeneous systems, that is, systems where there are
liquid interfaces such as dissolved solutes or ``hard walls.''
However, existing cDFTs do not accurately model the critical point of
a fluid.

It is the aim of renormalization group theory to tackle the critical
point. Calculations are difficult near the critical point because
there are large-scale fluctuations of the density. Indeed, there are
density fluctuations at \emph{all} length scales, which causes mean
field theories to fail. Renormalization group theory is an iterative
procedure that was developed in the 1950s, and Kenneth Wilson's work
with the theory in the 1970s won him the 1982 Nobel Prize in
physics.\cite{Nobel82} The end goal of this particular work is to
investigate into whether or not implementing renormalization group
theory is worthwhile to our research group. Many other groups use
renormalization group theory, but none of them report the iteration
depth used in their
calculations.\cite{Forte11,Ramana12,Ghobadi13,Wang13,Zhao13} If only
few iterations are required, then this could be an effective tool to
improve the accuracy of cDFTs. However, if many iterations are
required, the computational cost could be unwieldy.

In section~\ref{sec:theory}, I present some of the main theories I
deal with: thermodynamic perturbation
theory, the square well liquid free energy and renormalization group theory. In
section~\ref{sec:methods} I discuss the the methods used. I cover the
calculation of liquid-vapor coexistence curves in some detail, as
calculating coexistence near the critical point is tricky. The
results---such as they are---are covered in section~\ref{sec:results}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}\label{sec:theory}

@@

Thermodynamic perturbation theory (TPT) gives a good description of
simple liquids, as long a system is far from the critical point---which
is most of the time. TPT is a rigorous power series
expansion in a small parameter; as long as the parameter is small, the
theory is valid. This discussion follows primarily from \textit{Theory
  of Simple Liquids}, by Hansen and McDonald\cite{Hansen06}.

Thermodynamic perturbation theory treats the intermolecular potential
as separable into two parts, one for short-range repulsion and another
for long-range attraction. The short-range part is considered to be a
reference potential to which we add a perturbation. The following pair
potential $v(\III)$ describing the interaction between hard spheres is
our reference:
\begin{align}
  v_\text{hs}(\III) &=
    \begin{cases}
      \infty & r < \sigma \\
      0 & \sigma \leq r
    \end{cases}
\end{align}
To this reference, we take a square well attraction as a perturbation:
\begin{align}
  w_\text{sw}(\III) &= - \varepsilon \quad \sigma < r < \lambdaSW\sigma
\end{align}
so that we have
\begin{align}
  v_\xi(\III) &= v_\text{hs}(\III) + \xi w_\text{sw}(\III) \label{eqn:small-perturbation}
\end{align}
where $\xi$ is a parameter that allows us to make the perturbation arbitrarily small.
The total potential energy $\Vtilde$ of our system and its derivatives with
respect to $\xi$ (which become important later) are
\begin{align}
  \Vtilde &= \sum_{i=1}^N\sum_{j>i}^N v_\xi(\mathbf{r}_{ij})
\end{align}
\begin{align}
  \frac{\partial \Vtilde}{\partial\xi} &= \sum_{i=1}^N\sum_{j>i}^N w_\text{sw}(\mathbf{r}_{ij}) \nonumber \\
  &\equiv \Wtilde \\
  \frac{\partial^2 \Vtilde}{\partial\xi^2} &= 0
\end{align}

The \emph{configuration integral} (the potential energy component of
the partition function) and its derivative with respect to $\xi$
are
\begin{align}
  Z_\xi &= \int d\rr^N\, e^{-\beta \Vtilde}
\end{align}
\begin{align}
  \frac{\partial Z_\xi}{\partial\xi} &=  \int d\rr^N\, \frac{\partial}{\partial\xi}\left( e^{-\beta \Vtilde} \right) \nonumber \\
  &= -\int d\rr^N\, \beta\frac{\partial \Vtilde}{\partial\lambda}e^{-\beta \Vtilde} \nonumber \\
  &= -\int d\rr^N\, \beta \Wtilde e^{-\beta \Vtilde}
\end{align}
And the non-ideal part of the free energy (i.e.~the excess free
energy) and its derivative are
\begin{align}
  \Fex &= -\kT\ln\left( \frac{Z_\xi}{V^N} \right)
\end{align}
\begin{align}
  \frac{\partial \Fex}{\partial\xi} &= -\kT\frac{V^N}{Z_\xi}\frac{\partial}{\partial\xi}\left( \frac{Z_\xi}{V^N} \right) \nonumber \\
  &= -\kT\left( \frac{1}{Z_\xi}\frac{\partial Z_\xi}{\partial\xi} \right) \nonumber \\
  &= \frac{1}{Z_\xi} \int d\rr^N\, \Wtilde e^{-\beta \Vtilde} \label{eqn:dfdlambda_Z}\\
  &= \left\langle \Wtilde \right\rangle_{\xi} \label{eqn:dfdlambda}
\end{align}
(where $-\kT$ and $-\beta$ have divided out to $1$ before equation~\ref{eqn:dfdlambda_Z}). Here, $V^N$ is
the total volume taken to the $N^{th}$ power, and $\left\langle \Wtilde
\right\rangle_{\xi}$ is the ensemble average of $\Wtilde$, with the
subscript indicating some value of $\xi$.
Note that equation~\ref{eqn:dfdlambda} is exact; now take a series expansion around $\xi=0$:
\begin{align}
  \left\langle \Wtilde \right\rangle_{\xi} &= \left\langle \Wtilde\right\rangle_{\xi = 0} + (\xi)\frac{\partial\left\langle \Wtilde \right\rangle_{\xi}}{\partial\xi}\bigg|_{\xi = 0} + \mathcal{O}(\xi^2) \label{eqn:Wn-expansion}
\end{align}
The derivative term is
\begin{widetext}
  \begin{align}
    \frac{\partial}{\partial\xi}\left\langle \Wtilde \right\rangle_{\xi} &= \frac{\partial}{\partial\xi}\left( \frac{1}{Z_\xi}\int d\rr^N\, \Wtilde e^{-\beta \Vtilde}\right) \nonumber \\
    &= \frac{1}{Z_\xi}\frac{\partial}{\partial\xi}\left( \int d\rr^N\, \Wtilde e^{-\beta \Vtilde} \right) - \left( \int d\rr^N\, \Wtilde e^{-\beta \Vtilde} \right)\frac{1}{Z_\xi^2}\frac{\partial}{\partial\xi}Z_\xi \nonumber \\
    &= -\beta\left[ \frac{1}{Z_\xi}\int d\rr^N\, \Wtilde^2e^{-\beta \Vtilde} - \left( \frac{1}{Z_\xi}\int d\rr^N\, \Wtilde e^{-\beta \Vtilde} \right)^2 \right] \nonumber \\
    &= -\beta\left[ \left\langle \Wtilde^2 \right\rangle_{\xi} - \left\langle \Wtilde \right\rangle_{\xi}^2 \right] \nonumber \\
    &\equiv -\beta\sigma_{\xi}^2
  \end{align}
\end{widetext}
where $\sigma_xi^2$ is the variance of the perturbative potential \Wtilde. Combining this result with equation~\ref{eqn:Wn-expansion} gives us
\begin{align}
  \left\langle \Wtilde \right\rangle_{\xi} &= \left\langle \Wtilde\right\rangle_{\xi = 0} - \beta\xi\sigma_{\xi=0}^2 + \mathcal{O}(\xi^2) \label{eqn:Wn-expansion-simplified}
\end{align}
and we can insert this into equation~\ref{eqn:dfdlambda}:
\begin{align}
  \frac{\partial \Fex}{\partial\xi} &= \left\langle \Wtilde\right\rangle_{\xi = 0} - \beta\xi\sigma_{\xi=0}^2 + \mathcal{O}(\xi^2)
\end{align}
Now integrate this expression over $\Fex\in[F_\text{ref},\Fex]$ and $\xi\in[0,1]$ to obtain
\begin{align}
  \int_{F_\text{ref}}^{\Fex}\, d\Fex &= \int_0^1 d\xi\, \left[ \left\langle \Wtilde\right\rangle_{\xi = 0} - \beta\lambda\sigma_{\xi=0}^2 + \mathcal{O}(\xi^2) \right] \nonumber \\
  \hookrightarrow \Fex - F_\text{ref} &= (1-0)\left\langle \Wtilde \right\rangle_{\xi=0} - \frac{1}{2}\beta(1 - 0)^2\sigma_{\xi=0}^2 + \mathcal{O}(\beta^2) \nonumber \\
  \hookrightarrow \Fex &= F_\text{ref} + \left\langle \Wtilde \right\rangle_{\xi=0} - \frac{1}{2}\beta\sigma_{\xi=0}^2 + \mathcal{O}(\beta^2) \label{eqn:F-inTermsW} \\
  &\equiv F_\text{ref} + a_1(n) + a_2(n)
\end{align}
This is the \emph{high-temperature expansion}, first derived by
Zwanzig\cite{Zwanzig54}. He further showed that the $n^{th}$ term could be written in
terms of the mean fluctuations $\left\langle \left[ \Wtilde - \left\langle
  \Wtilde\right\rangle_{\xi=0} \right]^\nu \right\rangle_{\xi=0}$
with $\nu \leq n$. This requires us to know the form of higher-order
correlation functions of the reference system, which can be quite
complicated. In the next section we will see simple approximations for
the $n=1$ and $n=2$ terms.

\subsection{Square well liquid free energy}\label{subsec:SW}

As a baseline for RG, I use the square well liquid free energy:\cite{Hughes13}
\begin{align}
  f_\text{TPT}(T,n) &= \fid + \fhs + \fsw
\end{align}
$\fid$ is the ideal gas free energy, $\fhs$ is the free energy contribution due to hard sphere
repulsion, and $\fsw$ is the free energy contribution due to square well attraction. Each function is outlined below.
